### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md

> 1.      How relevant is this submission to the scope of JEDM? (if
>         applicable: to the targeted special issue?)
>
> The topic itself, categorizing subsequences of actions in a learning
> management system, is quite relevant.
>
>
> 2.      How novel is the described research? Are the authors aware of
>         related work?
>
> It is somewhat novel, but the authors do not seem to be adequately aware of
> related work in the methodology contribution. They seem to be even less
> aware of substantive work in education.

Thank you. We agree with your assessment of relevance to JEDM, and concede
that our knowledge of related work in education is a potential blind spot
for us as CS researchers.

> 3.      What is the scientific contribution of this submission? Is it
>         clearly explained, in terms of how the paper advances the EDM field
>         or contributes to related fields?
>
> This paper presents a novel method for learning categories of action
> sequences from MOOC clickstream logs. Although the authors have done a bit
> of background research on approaches to MOOC activity sequences, the
> educational content knowledge is very thin. The approach is under-theorized
> and does not adequately address the “why are you doing this” question.

We agree that we are weak on the educational aspect of the related work. We
have attempted to revise this based on your later comments. We also agree
that we may need to more clearly articulate our answer to the "why"
question from a more education-oriented perspective. Our original
description was perhaps too technically motivated and insufficiently
addresses this particular perspective. From education perspective, the motivation for our work is two-fold: 1) there is a need to understand student behavior in detail in the online education environment (which is challenging due to the scale of education), and the proposed model addresses this need (as an "intelligent assistant" for humans); 2) the naturally available student interaction data contains valuable knowledge/patterns about student behavior, which the proposed model can help reveal.  We hope our revised introduction
more clearly articulates why this is an important problem to tackle.

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md

> 3.      What is the scientific contribution of this submission? Is it
>         clearly explained, in terms of how the paper advances the EDM field
>         or contributes to related fields?
>
> This paper presents a novel method for learning categories of action
> sequences from MOOC clickstream logs. Although the authors have done a bit
> of background research on approaches to MOOC activity sequences, the
> educational content knowledge is very thin. The approach is under-theorized
> and does not adequately address the “why are you doing this” question.

We agree that we are weak on the educational aspect of the related work. We
have attempted to revise this based on your later comments. We also agree
that we may need to more clearly articulate our answer to the "why"
question from a more education-oriented perspective. Our original
description was perhaps too technically motivated and insufficiently
addresses this particular perspective. From education perspective, the motivation for our work is two-fold: 1) there is a need to understand student behavior in detail in the online education environment (which is challenging due to the scale of education), and the proposed model addresses this need (as an "intelligent assistant" for humans); 2) the naturally available student interaction data contains valuable knowledge/patterns about student behavior, which the proposed model can help reveal.  We hope our revised introduction
more clearly articulates why this is an important problem to tackle.

> The paper is on shaky ground because it attempts to do two things and
> does neither well. The first is develop a new method. The second is apply
> the method to learn about behavior patterns in MOOCs.

We respectfully disagree that we do neither well. We could have done a
better job of framing our work (which we feel we have improved in the
revision), but we feel that the model we use in this work is novel in the
education domain and that the behavior patterns we uncover are useful,
unique, and important. (No existing model can be directly applied to this data to produce the same output as the proposed model produces.)

@@

Specifically, we have now called out our model's relation to similar
two-layered approaches applied in other domains, articulated how our model
may be viewed as a special case of a more general Hierarchical Hidden
Markov Model with certain hard constraints, and discussed in more detail
how our model relates to the literature about mixtures of hidden Markov
models.

We have also attempted to also do a better job of discussing the potential
limitations and drawbacks of our method, including those related to the
application of the EM algorithm for fitting our model, in a new section
added to the paper.

> 4.      Is the work technically sound? Are there enough methodological
>         details? Are claims convincingly substantiated, either through
>         theoretical argument or empirical data?
>
> The current manuscript does not adequately develop the methodology for
> review on its technical soundness, and therefore the application is
> premature. Much of the substantive conclusions are the results of
> Rorschach-test-y looking at graphs.

While there are many results that do indeed require qualitative assessment
of latent state representations and transitions, we disagree that this
makes the method technically unsound. Our goal is to uncover patterns that
exist in this massive clickstream log data, which without a model are
prohibitively expensive for instructors to discover. We view our model as a
component in a human-machine collaborative system in which the
computational techniques employed by the machine are applied to do what
they are good at (namely, uncovering statistical patterns/trends) and the
human effort is leveraged to interpret the patterns and turn them into actionable knowledge.

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md

> 4.      Is the work technically sound? Are there enough methodological
>         details? Are claims convincingly substantiated, either through
>         theoretical argument or empirical data?
>
> The current manuscript does not adequately develop the methodology for
> review on its technical soundness, and therefore the application is
> premature. Much of the substantive conclusions are the results of
> Rorschach-test-y looking at graphs.

While there are many results that do indeed require qualitative assessment
of latent state representations and transitions, we disagree that this
makes the method technically unsound. Our goal is to uncover patterns that
exist in this massive clickstream log data, which without a model are
prohibitively expensive for instructors to discover. We view our model as a
component in a human-machine collaborative system in which the
computational techniques employed by the machine are applied to do what
they are good at (namely, uncovering statistical patterns/trends) and the
human effort is leveraged to interpret the patterns and turn them into actionable knowledge.

> There are selective statistical tests performed which violate some basic
> principles of statistical practice.

We feel that we cannot adequately address this criticism without a clear
articulation of exactly where in the paper such mistakes were made, and
what the proper practice should be in those cases. We would thus benefit greatly from more elaboration on this point.

@@

> The issue of temporal grain size/resolution seems important, but is very
> mysterious once the method is actually described. As it is sold, one would
> expect the model itself to determine when the (latent) state has switched.
> But it appears that the authors simply chose a cutoff to divide sequences
> into “one day’s worth of activity.” Why is that a reasonable choice? And
> why should tomorrow’s latent state depend on today’s state (which they do
> in a HMM)?

Choosing the cutoff points for what defines a browsing session is indeed
an important assumption that will impact the patterns that are uncovered.
However, rather than viewing this as a weakness, we can view this as a
strength in the sense that this decision depends entirely on the
granularity of the patterns you wish to discover (following our vision of building a software tool based on the proposed model for use as an intelligent assistant for humans).

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md

I will attempt to make an analogy with text documents. The space of all
possible English journal articles is unfathomably large, but we can still
model the generation of these articles using simpler statistical models. If
we have a discrete distribution over words, we can model the generation of
a document as repeatedly sampling the next word in the document from this
discrete distribution. This model is obviously naive and doesn't capture
everything we know about how documents are generated, but it is
"sufficient" in the sense that for any text document, the contents of it
can in fact be modeled by repeated sampling of words from some
distribution that covers its vocabulary. We do not require exponentially
many parameters to model document generation in this setting, and similarly
in this paper we do not require exponentially many parameters to model
sequence generation because we make a strong assumption about the kind of
parametric model that is used to generate the observed sequences.

> While the idea of breaking up a longer sequence of observed actions into
> shorter subsequences which are each characterized by a latent state is
> interesting, the usefulness of this approach in the present paper is less
> clear. The authors overstate the degree to which this method will help to
> automatically infer student behaviors, as much of the interpretability of
> the results still rests on human readings of the Markov chains.

We have attempted to modify the paper to be more precise about our claims
about what is automatically discovered and what is manually discovered. We
still feel, however, that this method is useful in the sense that these
patterns cannot be discovered from the raw clickstream data manually by
instructors without the application of a model that can provide statistical
summaries of the data such as the ones provided by our model, and no existing model can be used directly to generate the same behavior patterns as what our model generates.

### 2017-04-04 jedm-reviews/round-1/reviewer_c_response.md

> 8.      Are the presentation, organization and length satisfactory?
>
> The introductory section should be shortened, since the details on prior
> work are delayed for too long.

Thank you for the suggestion.

> 9.      Can you suggest additions or amendments or an introductory
>         statement that will increase the value of this paper?
>
> The method needs significantly more rigorous development.  Stylistically,
> the authors should try to be less boastful about the utility of this
> method, which is much more dubious than they make out.

We have attempted to be more clear in the introduction about exactly the
contributions we are making that are sufficiently supported by our existing
experiment results.

### 2017-04-05 jedm-reviews/round-1/reviewer_c_response.md

> The method itself is a twist on Markov switching models or on mixtures of
> Markov Chains and appears to be very related to the following works, which
> are not referenced:
>
> Hamilton, J. D. (1990). Analysis of time series subject to changes in
> regime. Journal of Econometrics, 45(1-2):39–70.
>
> Song, Y., Keromytis, A. D., & Stolfo, S. J. (2009, February). Spectrogram:
> A Mixture-of-Markov-Chains Model for Anomaly Detection in Web Traffic. In
> NDSS (Vol. 9, pp. 1-15).
>
> Gupta, R., Kumar, R., & Vassilvitskii, S. (2016). On Mixtures of Markov
> Chains. In Advances in Neural Information Processing Systems (pp.
> 3441-3449).
>
>
> The last reference is admittedly recent and used spectral methods. However,
> the important point is that these above references all offer more
> systematic descriptions of model properties, for example by studying the
> recoverability of parameters in a simulation study. (Note that the Gupta et
> al. reference showed that EM estimation was error-prone and needed to be
> run on the order of 1000 times to recover parameters as well as the
> spectral approach.)

We agree that our related work section was weak in the breadth sense and
perhaps focuses too much on MOOC-specific or CS-specific research and did
not cover enough related work in wider domains. We wholeheartedly thank you
for your suggested references, which we have incorporated into a
substantially revised related work section that more clearly articulates
how our model differs from similar models that have been applied in other
domains.

Specifically, we have now called out our model's relation to similar
two-layered approaches applied in other domains, articulated how our model
may be viewed as a special case of a more general Hierarchical Hidden
Markov Model with certain hard constraints, and discussed in more detail
how our model relates to the literature about mixtures of hidden Markov
models.

We have also attempted to do a better job of discussing the potential
limitations and drawbacks of our method, including those related to the
application of the EM algorithm for fitting our model, in a new section
added to the paper.

### 2017-04-05 sec-conclusion.tex
Several minor typo fixes.

Although we only experimented with our model on two MOOCs, the model
is completely general and can be easily applied to analyze the log of any other
course to enable deep understanding of student behaviors as well as the coorrelations
of such behaviors as other variables such as grades. We plan to develop a MOOC log analysis
system based on the proposed model to both facilitate education research and 
help instructors improve course design.

>>>>

Although we only experimented with our model on two MOOCs, the model
is completely general and can be easily applied to analyze the log of any other
course to enable deep understanding of student behaviors as well as the correlations
of such behaviors and other variables such as grades. We plan to develop a MOOC log analysis
system based on the proposed model to both facilitate education research and 
help instructors improve course design.

### 2017-04-05 sec-limitations.tex
Several minor typo fixes.

Our model fits its parameters using maximum likelihood estimation using the
EM algorithm. The EM algorithm is a hill climbing algorithm that is
optimizing in a highly non-convex parameter space. Thus, it can only
guarantee reaching a local maximum in practice~\citep{Dempster:1977:JRSS}.
This may mean that the parameters found for the model may not be the
``best'' parameters in a global sense, which may lead to suboptimal latent
state representations and transition patterns. A commonly applied approach
to address this drawback is to run the model multiple times with different
starting points to allow the model to explore a larger portion of the
potential parameter space. One can then compare the log-likelihood of the
data between the models that were started from different initialization
points and select the one that has the highest value. This still does not
guarantee finding a global maximum, but it does help alleviate the
potential for finding a particularly bad local maximum. In practice, we
have found that our model model can be fit to the data fairly quickly on
commodity hardware, so running it multiple times to address this concern is
not computationally unfeasible.

A further complication in blindly applying the EM algorithm to our model is
the fact that the observation probabilities will be incredibly small. The
probability that a specific sequence is generated by a specific Markov
chain (one of our latent states) will decrease exponentially with its
length. While there are general approaches to avoiding numerical underflow
in hidden Markov models, applying the ``scaling'' method proposed by
\citet{Rabiner:1990:RSR} will still result in numeric stability issues in
our case due to the incredibly small sequence-generation probabilities. We
address this in our open-source implementation by computing the trellises
in log-space and using the log-sum-exp trick when we need to take
summations, which exploits the identify
\begin{equation}
    \log \sum_{i=1}^n e^{x_i} = a + \log \sum_{i=1}^n e^{x_i - a}
\end{equation}
where $a$ is typically set to $\max_i x_i$ to improve stability. We did not
encounter further stability issues once we applied these two tricks.

>>>>

Our model fits its parameters using maximum likelihood estimation using the
EM algorithm. The EM algorithm is a hill climbing algorithm that is
optimizing in a highly non-convex parameter space. Thus, it can only
guarantee reaching a local maximum in practice~\citep{Dempster:1977:JRSS}.
This may mean that the parameters found for the model may not be the
``best'' parameters in a global sense, which may lead to suboptimal latent
state representations and transition patterns. A commonly applied approach
to address this drawback is to run the model multiple times with different
starting points to allow the model to explore a larger portion of the
potential parameter space. One can then compare the log-likelihood of the
data between the models that were started from different initialization
points and select the one that has the highest value. This still does not
guarantee finding a global maximum, but it does help alleviate the
potential for finding a particularly bad local maximum. In practice, we
have found that our model can be fit to the data fairly quickly on
commodity hardware, so running it multiple times to address this concern is
not computationally unfeasible.

A further complication in blindly applying the EM algorithm to our model is
the fact that the observation probabilities will be incredibly small. The
probability that a specific sequence is generated by a specific Markov
chain (one of our latent states) will decrease exponentially with its
length. While there are general approaches to avoiding numerical underflow
in hidden Markov models, applying the ``scaling'' method proposed by
\citet{Rabiner:1990:RSR} will still result in numeric stability issues in
our case due to the incredibly small sequence-generation probabilities. We
address this in our open-source implementation by computing the trellises
in log-space and using the log-sum-exp trick when we need to take
summations, which exploits the identity
\begin{equation}
    \log \sum_{i=1}^n e^{x_i} = a + \log \sum_{i=1}^n e^{x_i - a}
\end{equation}
where $a$ is typically set to $\max_i x_i$ to improve stability. We did not
encounter further stability issues once we applied these two tricks.

### 2017-04-05 sec-model.tex
Several minor typo fixes.

Given a MOOC log, we can define a set $\mathbf{A}$ of actions that a
student can take at any given time. For example, an action $a \in
\mathbf{A}$ might be ``viewing lecture'', ``taking quiz'', or ``viewing
forum''. For each student in the course $\ell \in \mathbf{L}$, we then
extract a list of action sequences $\mathbf{S}_\ell$ that he or she
produced as observed in the log, where each sequence $\mathbf{s} \in
\mathbf{S}_\ell$ is itself a list of actions $(a_1, a_2, \ldots, a_M)$ with
each $a_i \in \mathbf{A}$.  Each sequence can be divided flexibly; in this
paper we chose to denote the end of a sequence as occurring when no further
actions occur within a 10 hour window of time (and thus our sequences
roughly correspond to one day's worth of activity). This decision reflects
our desire to uncover latent state transition behavior at the granularity
of day-to-day behavior. A different segmentation strategy could be used to
uncover hour-by-hour behavior, or week-by-week behavior, and this depends
entirely on the desired time resolution one wishes to be exhibited in the
transitions between the latent states. Different segmentation strategies
will result in different underlying training data, and this different
meanings behind the patterns that the 2L-HMM will extract.

>>>>

Given a MOOC log, we can define a set $\mathbf{A}$ of actions that a
student can take at any given time. For example, an action $a \in
\mathbf{A}$ might be ``viewing lecture'', ``taking quiz'', or ``viewing
forum''. For each student in the course $\ell \in \mathbf{L}$, we then
extract a list of action sequences $\mathbf{S}_\ell$ that he or she
produced as observed in the log, where each sequence $\mathbf{s} \in
\mathbf{S}_\ell$ is itself a list of actions $(a_1, a_2, \ldots, a_M)$ with
each $a_i \in \mathbf{A}$.  Each sequence can be divided flexibly; in this
paper we chose to denote the end of a sequence as occurring when no further
actions occur within a 10 hour window of time (and thus our sequences
roughly correspond to one day's worth of activity). This decision reflects
our desire to uncover latent state transition behavior at the granularity
of day-to-day behavior. A different segmentation strategy could be used to
uncover hour-by-hour behavior, or week-by-week behavior, and this depends
entirely on the desired time resolution one wishes to be exhibited in the
transitions between the latent states. Different segmentation strategies
will result in different underlying training data, and thus different
meanings behind the patterns that the 2L-HMM will extract.

### 2017-04-05 sec-related-work.tex
Several minor typo fixes.

For example, \citet{Kizilcec:2013:LAK} assigned students to states
following a rule-based approach based upon when the student submitted the
assignment for a particular week in the course. They investigated how
students transitioned between these states as the course progressed, and
used the sequence of states a student exhibited as a representation for
performing $k$-means clustering of students into related groups. This
differs from our method substantially: we assign students to states using a
probabilistic framework to account for uncertainty in this state
assignment, and jointly learn \emph{representations} for these states,
which are treated as being \emph{latent} as opposed to pre-definied using
some rule (or set of rules).  Furthermore, our model provides more
flexibility in how the time segments are defined, allowing for both finer
(for example, day-by-day) or courser (for example, month-by-month)
granularity. \citet{Shih:2010:EDM} investigated the use of HMM-based
clustering techniques for automatic discovery of student learning
strategies when solving a particular problem.  This is similar to our
approach in that the description of behavior profiles is a Markov model,
but cannot further characterize each latent state with another informative
HMM. Thus their work can be regarded to modeling ``micro'' behavior,
whereas our model can model both ``micro'' and ``macro'' behavior.

>>>>

For example, \citet{Kizilcec:2013:LAK} assigned students to states
following a rule-based approach based upon when the student submitted the
assignment for a particular week in the course. They investigated how
students transitioned between these states as the course progressed, and
used the sequence of states a student exhibited as a representation for
performing $k$-means clustering of students into related groups. This
differs from our method substantially: we assign students to states using a
probabilistic framework to account for uncertainty in this state
assignment, and jointly learn \emph{representations} for these states,
which are treated as being \emph{latent} as opposed to pre-defined using
some rule (or set of rules).  Furthermore, our model provides more
flexibility in how the time segments are defined, allowing for both finer
(for example, day-by-day) or coarser (for example, month-by-month)
granularity. \citet{Shih:2010:EDM} investigated the use of HMM-based
clustering techniques for automatic discovery of student learning
strategies when solving a particular problem.  This is similar to our
approach in that the description of behavior profiles is a Markov model,
but cannot further characterize each latent state with another informative
HMM. Thus their work can be regarded to modeling ``micro'' behavior,
whereas our model can model both ``micro'' and ``macro'' behavior.
